A Survey of Physical Design Techniques of Information Systems

### 2010-07-24 cs848-db/Paper.tex

%\documentclass[conference]{IEEEtran}
\documentclass[12pt,a4paper]{article}
\usepackage{fullpage}

\usepackage{graphics}
\usepackage{color}
\usepackage[pdftex]{hyperref}
\usepackage{graphicx}
\usepackage{amssymb}

@@

Relational Database Systems have been the most popular type of Information Systems for decades. When we think of designing a database systems, we usually think
of which data we need to keep track of and how will we properly organize it into meaningful tables. However, this is not the only phase in database design.
All of our designed tables need to be physically stored on disk. How these tables will be stored and accessed greatly affects the database performance. Physical
database design is concerned with the actual storage of data on disk. This means how the files are organized, and how they are accessed. Data is stored in the
form of records in files which reside on a physical disk. In this sense, we try to find the most efficient way possible to store and access the data such that
queries are answered and executed efficiently. In the traditional relational database system, this involved selecting proper indexing techniques,
partitioning the data for more efficient access, materializing the results of some queries, clustering and compressing the data, and many other
techniques~\cite{fikelstein1988physical,lightstone2007physical}. The physical design aspects of relational databases have been studied in much detail. However,
over time, different types of Information Systems have emerged that have different characteristics. For example, most relational databases have been disk-based
where the primary copy of the data resides on disk. This is as opposed to Main Memory Relational Databases (MMDB) where the primary copy of the data resides in
main memory and no disk operations are needed to query the data. Such differences pose new requirements for the physical design of these systems.

This survey presents the different physical design patterns that have been used in different types of information systems. The systems discussed in this survey
are disk based relational databases (traditional database systems), main memory relational databases, data warehouses, and XML databases. The aim of this survey
is not to explain the details of every data structure used in physical design, but rather to compare how the different data structures have been used in these
systems. This includes explaining how alterations or additions are made to data structures to fit the specific needs of every system. For a complete reference
on the details of the data structures used in physical database design, please see (CITE HERE). Using the different physical design methods used in disk based
relational database systems, we explore how these methods apply to the other systems and where the differences and similarities lie.

The rest of the survey is organized as follows. Section~\ref{SEC-ELEMENTS} first explains different elements of physical database design based on traditional
disk-based databases. Section~\ref{SEC-DIFFSYS} then discusses how each of the three other systems uses these elements, as applicable, to improve its
efficiency.Sections~\ref{SEC-MMDB},~\ref{SEC-XML},~\ref{SEC-WAREHOUSES}, discuss Main Memory Databases, XML Databases, and Data Warehouses
respectively. Throughout these sections, comparisons are made as to how the different elements have been modified to fit the needs of each system.
Section~\ref{SEC-AUTO} then presents how the choice of the different physical design techniques can be automated. Section~\ref{SEC-OPEN} mentions some of the
open problems of physical database design, and suggests possible solutions. Finally Section~\ref{SEC-CONCL} summarizes and concludes this survey.

@@

Partitioning is an important aspect of physical design as it reduces table scanning time. There are two main categories of partitioning: horizontal partitioning
and vertical partitioning. Horizontal partitioning divides the table into sets of table rows where each row or record still has all the attributes of the table.
For example, dividing the data by date where all data dating less than year 2000 lies in one partition while all data dating more lies in another. On the other
hand, vertical partitioning reduces the width of the dataset by storing some attributes in one table, and some in another.
The main advantage of any type of partitioning is to reduce the amount of time it takes to scan a table which in turn improves performance. For example, if a
table has 1000 records, and a query is only using the first 20 records then having these 20 records in a partition containing 50 records will save time as
opposed to examining all the 1000 records. Further classification in partitioning is single vs group horizontal or vertical
partitioning~\cite{gruenwald1990database}. Horizontal partitioning just divides the data into partitions where each group contains complete tuples. Group
horizontal partitioning groups tuples that are more frequently used together. Single vertical partitioning divides a relation into groups each of which contains
only one type of attribute. Group vertical partitioning divides the data vertically such that attributes that are commonly required together are physically
stored together.

Put table from database about the diff horizontal partitioning 
MySQL only supports horizontal partitioning.

\subsection{Clustering}
When only records satisfying certain criteria are need from a table, it is often unnecessary to scan the whole table. For example, if only records that occurred
in January 2010 are needed, then querying all the table would be an extremely unnecessary overhead. This is where clustering comes in. Clustering reflects how
the records are physically located together on disk. Records that are likely to be queried together will be stored physically together to avoid querying the
whole table. For the previously mentioned example, the data can be clustered by month, such that all the records of January 2010 are physically in the same
page, for example. In this case, only this page will need to be read. As
people realized the potential of clustering, multidimensional clustering was introduced where data is clustered based on different criteria at the same time
(TALK MORE).

@@

A Main Memory Database (MMDB) (sometimes referred to as in-memory database) is one where the data resides in the main memory of the system rather than on a
disk~\cite{garcia1992main}. It should be noted that this is different from caching. Disk based database systems use the main memory to cache query results.
However, MMDB's primary copy of the data resides in main memory. MMDB have many implementation challenges that have been addressed throughout the research
community.

Garcia-Molina and Salem~\cite{garcia1992main} mention some of the challenges involved. One is concurrency control. In disk based databases, locks are tracked
through a hash table, but the actual objects on disk do not contain lock information. On the other hand, in MMDB, lock status is part of the object itself since
it is cheap to keep a number of bits for that. The next challenge is access methods. In disk-based databases, B-Trees are used to index the data for faster
access. B-Trees have a short bushy nature to try to decrease the height of the tree to decrease the number of I/O accesses. Since I/O access is not a problem in
MMDB, longer tree structures are used since its cheap to access main memory. Another point is that data values do not need to be stored in the index itself
since they will be stored in main memory anyways so there are no performance gains from storing them in the index itself. Table~\ref{tab:mmdbsumm} summarizes
the physical design elements of MMDBs.

\begin{table}[t!]
\centering
\scalebox{0.7}{
\begin{tabular}{p{3cm}|p{6cm}|p{6cm}}
\hline
 \textbf{Physical Design Element} & \textbf{Desired Features} & \textbf{Methods/Structures Used}\\
\hline\hline
Index Structures & \begin{itemize}
                    \item No need to store actual values in index \item Larger node size that is aligned with cache line size for cache consciousness
                   \end{itemize} & \begin{itemize}
                                    \item B Tree
\item B+ Tree
\item CSB+ Tree
\item pB+ Tree
\item T Tree
\item CST Tree
\item CSS Tree
                                   \end{itemize}\\
\hline
Materialized Views & \begin{itemize} \item Not needed since processing and memory access is cheap \end{itemize}& N/A\\
\hline
Partitioning & \begin{itemize} \item Only needed in secondary storage to speed up reloading in case of a crash \end{itemize} & \begin{itemize}
                                                                                            \item Horizontal Partitioning
\item Single Vertical Partitioning
                                                                                           \end{itemize}\\
\hline
Clustering & \begin{itemize} \item Not needed since sequential random or dispersed main memory access is not more expensive than sequential access \end{itemize}
& N/A

@@

Two considerations are taken when designing index structures for MMDB. The first is that the data resides in main memory and not on disk, and so many of the
considerations taken for I/O operations is no longer there. Instead of focusing on disk access and disk storage, the data structures for main memory
databases should focus on the efficient use of CPU cycles and memory~\cite{lehman1986study}. The second is having index structures cache conscious. Cache
memories improve performance by holding recently referenced data~\cite{smith1982cache}. If the memory reference is found in the cache, then execution proceeds
at processor speed. Otherwise, the data has to be fetched from main memory. With the big difference between processor speed and main memory speed, cache misses
are still relatively expensive. Lots of research work has shown that cache performance is very critical in MMDBs~\cite{boncz1999database,rao1999cache}. This is
because the big difference between processor speed and main memory access speed. Thus, a cache miss is still relatively expensive in main memory databases.
Accordingly, we will see that many of the index data structures proposed for main memory databases focused on being cache conscious or cache sensitive. In
summary, a main memory index should be able to reduce overall computation time while using as little memory as possible~\cite{lehman1986study}. It should also
be noted that unlike disk based systems where it is advantageous to store actual attribute values in the index, main memory systems do not require that.
Instead, pointers to the actual attribute values can be used. This, in turn, will reduce the size of the index which is preferable in this case.

In this section, we talk about three families of index structures used in MMDB: B Trees, T Trees, and Binary Search Trees. There are more recent index
structures developed for MMDB such as J+ Trees~\cite{luan2009prefetching}, but which have not gained much popularity, and so we do not discuss them in our
survey. Table~\ref{tab:mmdbindexsumm} summarizes the features of the different index structures discussed in this section with respect to MMDB. The T Tree is
the most used index for MMDB (E.g. MySQL CLuster). However, there are other indexes used as well. For example, IBM's solidDB uses Trie (or prefix tree) for its
indexing. For their purposes, it served as a good index for main memory since it eliminates the need for many key comparisons~\cite{ibmsoliddb}.

\begin{table}[t!]
\centering
\scalebox{0.7}{
\begin{tabular}{p{2cm}|p{6cm}|p{6cm}}
\hline
 \textbf{Index Structure} & \textbf{Features suitable for MMDB} & \textbf{Cache Consciousness}\\
\hline\hline
B Trees &
\begin{itemize}\item Good storage utilization
\item Reasonably quick searching
\item Fast updating
\end{itemize} 
& \begin{itemize}\item Reasonable cache behavior if node fits in cache line\end{itemize}\\
\hline
B+ Trees & 
\begin{itemize}
\item Wastes space since all data is stored in the leaves
\item Reasonably quick searching
\item Fast updating
\end{itemize} &
\begin{itemize}
\item Reasonable cache behavior if node fits in cache line\end{itemize}\\
\hline
CSB+ Trees & \begin{itemize}
              \item Same features of B+ Trees
             \end{itemize} &
\begin{itemize}
 \item Good cache performance due to improved locality \& lower tree height
\end{itemize}\\
\hline

pB+ Trees & \begin{itemize}
              \item Same features of B+ Trees
             \end{itemize} &
\begin{itemize}
 \item Improves cache miss performance by prefetching more data in each cache miss
\item Node size is also increased leading to the same benefits of CSB+ Trees
\end{itemize}\\
\hline
T Tree &  \begin{itemize}
              \item Contain pointers to data values instead of the values themselves leading to better space utilization
\item Node size is bigger leading to previous mentioned advantages
             \end{itemize} &
\begin{itemize}
 \item Cache behavior was not considered at time of design
\end{itemize}\\
\hline
CST & \begin{itemize}
       \item Binary search tree for maximum value of each node leads to faster search
\item More space efficient by removing pointers \& using array storage. Index calculation used to allocate values
      \end{itemize} &
\begin{itemize}
 \item Node size is aligned with cache line size to avoid misses
\end{itemize}\\
\hline
CSS Trees &  \begin{itemize}
       \item Fast traversal in $log_{m+1}n$ ($m$ is the number of keys per node)
\item Mainly suitable for read environments
      \end{itemize} &
\begin{itemize}
 \item Good cache behavior since $m$ is chosen to fit in the cache line size
\end{itemize}\\

@@

T Trees were the main index structure used for main memory databases for some time after their proposal. However, it was discovered that B+-trees outperform
T-Trees on modern processors because of the growth of cache miss latency relative to processor speed~\cite{rao1999cache,lee2007cst}. This is because the height
of the tree is high which makes the total number of memory accesses from the root to the leaf node higher. The other reason is that the node size is not aligned
with the cache line size. Another problem with T-Trees is that there is a big waste of space in the cache because unnecessary data is brought into the cache.
Additionally, record pointers stored in the tree take up a lot of space.

To resolve these issues, Lee et al.~\cite{lee2007cst} propose the Cache Sensitive T-Trees (CST-Trees). This is achieved in the following ways. First, a binary
search tree whose node values are the maximum key of each node in the T Tree is constructed. That way, for a given value, the node containing this value could
be quickly located through searching the binary tree. Second, the need to store pointers is eliminated by storing the tree in an array and locating the
necessary nodes through index calculation. Finally, node sizes are aligned with cache line size such that there are no cache misses when accessing data in each
binary search tree in the array.

@@

Binary Search Trees have poor cache behavior when the array is much bigger than the cache. Rao and Ross~\cite{rao1999cache} propose the Cache Sensitive Search
Tree (CSS-Tree) to improve this cache behavior. To do that, the search tree is stored as a directory in an array. The number of keys per node is chosen such
that the whole node fits in a cache line. This improves local searching within a node to just one cache miss. They also hard-code the traversal within a node
such that finding the next node happens in $log_{m+1}n$ times rather than $log_{2}n$ times in a binary search tree where $m$ is the number of keys per node.
However, CSS Trees are only suitable for read environments since they eliminate nearly all child pointers, thus removing the support for incremental update.

\subsubsection{Materialized Views}
From our literature survey, it seems that materialized views are not used in MMDB. This makes sense since the problems materialized views try to address are
not present in MMDB. That is, the cost of computing complicated join queries is much less in MMDB due to the low cost of accessing the data in memory.
Accordingly, the maintenance cost of materialized views in MMDB will outweigh their minimal benefit.

@@

Partitioning is used in disk based systems to divide the data into the physical pages in a way that will yield the best performance. This could be through
vertical partitioning or horizontal partitioning, as previously explained. In main memory databases, we do not have the issue of expensive disk access which
makes partitioning in memory unnecessary as it will not save any costs. However, partitioning is needed in the secondary storage of the data on
disk~\cite{gruenwald1990database}. Disk storage is still used as a backup of main memory databases. When a crash occurs, a reload of the database from disk or
archive memory takes place. In order to avoid page faults caused by referencing data that is still being reloaded, the reloading process must be designed to be
efficient in loading the important data such that the database users are not affected. An efficient MMDB reload takes into consideration the number of I/Os for
reload and the number of main memory references during transaction processing. Gruenwald and Eich~\cite{gruenwald1990database,gruenwald1990choosing} show that
horizontal partitioning is the most suitable when the database performs more modifications than tuple deletions and more selections that projections and joins.
Single vertical partitioning is chosen in the other cases. They also show that if we just concentrate on the reload performance (i.e the number of I/Os for
reload), then single vertical partitioning is always the best choice.

@@

XML (eXtensible Markup Language) is increasingly being used as a data exchange format. Accordingly, the ability to store and manage XML documents is needed.
Salminen and Tompa~\cite{salminen2001requirements} define an XML database as ``a collection of XML documents and their parts, maintained by a system
having capabilities to manage and control the collection itself and the information represented by that collection.'' Salminena and Tompa also highlight some of
the requirements of an XML database system. These requirements include dealing with namespaces, Internet resources, querying parts of an XML document, and
transformations of XML documents.

There are two ways in which an XML database system can be designed. The first is to have the database internally translate the XML document into relational
tables (XML-enabled databases). The second is to have data structures that can persistently store XML documents (native XML databases). In this section, we will
discuss how the database's physical design needs to be adapted in both cases. Other challenges such as expressing and executing queries and updates are faced in
XML database design, but these are beyond the scope of this survey.

%eXist~\cite{meier2009exist} and TIMBER~\cite{jagadish2002timber} are two of the famous research efforts to design a native XML database.

@@

Despite the special nature of XML documents, the same index structures used in relational databases can still be utilized in XML databases. However, some
adjustments may need to be made to fit the nature of the XML documents. For XML-enabled databases, XML documents are usually represented as relational tables,
and then indexed similar to other tables. Pal et al.~\cite{pal2004indexing} describe how this is done in Microsoft SQL Server 2005. To start with, a new data
type called `XML' was introduced. This data type could contain values of complete XML documents, or just fragments of XML data. First, the different nodes in
the XML data are labelled with the ORDPATH~\cite{o2004ordpaths} mechanism. The XML data is then shredded into a relational table with five main columns:ORDPATH,
TAG, NODE\_TYPE, VALUE, and PATH\_ID. This information is then stored in a $B^{+}-tree$ as the primary index of the XML type. Additionally, a secondary index
can be created on any of the columns of the primary index to speed things up.

For native XML databases such as eXist~\cite{meier2009exist} or TIMBER~\cite{jagadish2002timber}, $B^{+}-trees$ were still used to index the XML documents.
However, in contrast to XML-enabled databases, the XML data is not first stripped into a relational table. eXist uses a number schema to assign unique
identifiers to each node in the XML document. These unique number identifiers allow the determination of any node's parent, sibling or possible child nodes.
Four index files are created for XML data where all the indexes are based on $B^{+}-trees$. One index manages the collection hierarchy, one index collects nodes
in a paged file and associates unique node identifiers to the actual nodes, one index indexes elements and attributes, and the last index keeps track of word
occurrences and is used by the fulltext search extensions. Oracle uses a similar technique for indexing its XML content where an XMLIndex table is created for
XML entries~\cite{oracleXML}. This table contains a unique identifier for the XPath leading to the node, the row id of the table used to store the XML data,an
order key that identifies the hierarchal positioning of the node, a locator key used for fragment extraction, and the text value of the node.

>>>>

\documentclass[12pt,a4paper]{article}
\usepackage{fullpage}

\usepackage{graphics}
\usepackage{color}
\usepackage[pdftex]{hyperref}
\usepackage{graphicx}
\usepackage{amssymb}
\usepackage{url}
\usepackage{multirow}

@@

Relational Database Management Systems (RDBMS) have been the most popular type of Information Systems for decades. When we think of designing a database
system, we usually think of which data we need to keep track of and how will we properly organize it into meaningful tables. However, this is not the only phase
in database design. All of the designed tables need to be physically stored on disk. How these tables will be stored and accessed greatly affects the database
performance. Physical database design is concerned with the actual storage of data on disk. This means how the files are organized, and how they are accessed.
Data is stored in the
form of records in files which reside on a physical disk. In this sense, we try to find the most efficient way possible to store and access the data such that
queries are answered and executed efficiently. In the traditional relational database system, this involved selecting proper indexing techniques,
partitioning the data for more efficient access, materializing the results of some queries, clustering and compressing the data, and many other
techniques~\cite{finkelstein1988physical,lightstone2007physical}. The physical design aspects of relational databases have been studied in much detail. However,
over time, different types of Information Systems have emerged that have different characteristics. For example, most relational databases have been disk-based
where the primary copy of the data resides on disk. This is as opposed to Main Memory Relational Databases (MMDB) where the primary copy of the data resides in
main memory and no disk operations are needed to query the data. Such differences pose new requirements for the physical design of these systems.

This survey presents the different physical design patterns that have been used in different types of information systems. The systems discussed in this survey
are disk based relational databases (traditional database systems), main memory relational databases, data warehouses, and XML databases. The aim of this survey
is not to explain the details of every data structure used in physical design, but rather to compare how the different data structures have been used in these
systems. This includes explaining how alterations or additions are made to data structures to fit the specific needs of every system. Using the different
physical design methods used in disk based
RDBMSs, we explore how these methods apply to the other systems and where the differences and similarities lie.

The rest of the survey is organized as follows. Section~\ref{SEC-ELEMENTS} first explains different elements of physical database design used in traditional
disk-based databases. Section~\ref{SEC-DIFFSYS} then discusses how each of the three other systems uses these elements, as
applicable. Sections~\ref{SEC-MMDB},~\ref{SEC-XML},~\ref{SEC-WAREHOUSES}, discuss Main Memory Databases, XML Databases, and Data Warehouses
respectively. Throughout these sections, comparisons are made as to how the different elements have been modified to fit the needs of each system.
Section~\ref{SEC-OPEN} mentions some of the current
open problems of physical database design. Finally Section~\ref{SEC-CONCL} summarizes and concludes this survey.

@@

Partitioning is an important aspect of physical design as it reduces table scanning time~\cite{lightstone2007physical}. There are two main categories of
partitioning: horizontal partitioning
and vertical partitioning. Horizontal partitioning divides the table into sets of table rows where each row or record still has all the attributes of the table.
For example, dividing the data by date where all data dating less than year 2000 lies in one partition while all data dating more lies in another. On the other
hand, vertical partitioning reduces the width of the dataset by storing some attributes in one table, and some in another. Usually, less frequently queried
columns are partitioned together to reduce the size of the table.

The main advantage of any type of partitioning is to reduce the amount of time it takes to scan a table which in turn improves performance. For example, if a
table has 1000 records, and a query is only using the first 20 records then having these 20 records in a partition containing 50 records will save time as
opposed to examining all the 1000 records. Horizontal partitioning is more widely used, and is implemented in all current DBMSes.

%Table~\ref{tab:horpartition}
shows different categories of horizontal partitioning, and the situations they are suited to. Combining partitioning and indexing can often lead to better
performance. For example, if List partitioning is used to divide a table by regions and then a bitmap index is used on that table, finding records in the same
region or state can be much faster. Finally, partitioning the data into different physical devices to achieve parallelism when accessing it can greatly
increase the performance of the database.

NEED SOME REFERENCES HERE

\begin{table}[!t]
\scalebox{0.7}{
 \begin{tabular}{p{2cm}||p{4cm}|p{4cm}|p{4cm}|p{4cm}}
& \textbf{Range Partitioning} & \textbf{List Partitioning} & \textbf{Hash Partitioning} & \textbf{Composite Partitioning}\\
\hline\hline
\textbf{How?} & If key falls in range & List of keys & Hash function& Hybrid\\
\hline
\textbf{Why?} & Disk limits & Group point values & Range/List are N/A & According to needs\\
\hline
\textbf{Benefits} & \multicolumn{4}{c}{Partition elimination, Improved administration, Fast roll-in/out, and Implicit clustering }\\
\hline
\textbf{Example} & \begin{itemize}
                    \item Zip code 
		    \item Time range
                   \end{itemize} &
		    \begin{itemize}
		    \item Regions
		    \item States
		    \end{itemize} &
		    \begin{itemize}
		    \item EmployeeId
		    \item ProductId
		    \end{itemize} &
		    Hash then List \\
\hline
\textbf{RDBMS} & MySQL, DB2, Oracle, PostgreSQL, SQL Server & MySQL, Oracle, PostgreSQL, SQL Server & MySQL, Oracle & MySQL, Oracle\\
\end{tabular}
}
\caption{Types of horizontal partitioning}
\label{tab:horpartition}
\end{table}

\subsection{Clustering}
Clustering reflects how the records are physically located together on disk. This is very important since sequential I/O access is much cheaper than random
access. Therefore, records that are likely to be queried together should be stored physically together to decrease the number of random I/O accesses.
MOREEEEEEEEEEEEEEEE

@@

A Main Memory Database (MMDB) (sometimes referred to as in-memory database) is one where the primary copy of the data resides in the main memory of the system
rather than on a
disk~\cite{garcia1992main}. It should be noted that this is different from caching. Disk based database systems use the main memory to cache query results.
However, MMDB's primary copy of the data resides in main memory. Accessing data in main memory is faster than disk, and so MMDBs usually have better
performance that disk based systems. However, main memory is more expensive and the size is limited in comparison to disk. However, MMDBs are essential to use
where some of the collected data cannot, or will not, be stored on disk in the first place. For example, phone switch data may be needed for reasoning with
other stored data, but it is captured in real time and will not necessarily be stored on disk. MMDB have many implementation challenges that have been addressed
throughout the research community. These challenges will be discussed throughout this section. %Table~\ref{tab:mmdbsumm} summarizes
the physical design elements of MMDBs.

\begin{table}[t!]
\centering
\scalebox{0.9}{
\begin{tabular}{p{3cm}|p{6cm}|p{6cm}}
\hline
 \textbf{Physical Design Element} & \textbf{Desired Features} & \textbf{Methods/Structures Used}\\
\hline\hline
Index Structures & \begin{itemize}
                    \item No need to store actual values in index \item Larger node size that is aligned with cache line size for cache consciousness
                   \end{itemize} & \begin{itemize}
                                    \item B Tree
\item B+ Tree
\item CSB+ Tree
\item pB+ Tree
\item T Tree
\item CST Tree
\item CSS Tree
                                   \end{itemize}\\
\hline
Materialized Views & \begin{itemize} \item Not highly utilized since processing and memory access is cheap \end{itemize}& N/A\\
\hline
Partitioning & \begin{itemize} \item Only needed in secondary storage to speed up reloading in case of a crash \end{itemize} & \begin{itemize}
                                                                                            \item Horizontal Partitioning
\item Vertical Partitioning
                                                                                           \end{itemize}\\
\hline
Clustering & \begin{itemize} \item Not needed since random or dispersed main memory access is not more expensive than sequential access \end{itemize}
& N/A

@@

Garcia-Molina and Salem~\cite{garcia1992main} mention data access methods as one of the main challenges involved in MMDBs. Two considerations are taken when
designing index structures for MMDB. The first is that the data resides in main memory and not on disk, and so many of the
considerations taken for I/O operations is no longer there. Instead of focusing on disk access and disk storage, the data structures for main memory
databases should focus on the efficient use of CPU cycles and memory~\cite{lehman1986study}. The second is having index structures cache conscious. Cache
memories improve performance by holding recently referenced data~\cite{smith1982cache}. If the memory reference is found in the cache, then execution proceeds
at processor speed. Otherwise, the data has to be fetched from main memory. With the big difference between processor speed and main memory speed, cache misses
are still relatively expensive. Lots of research work has shown that cache performance is very critical in MMDBs~\cite{boncz1999database,rao1999cache}. This is
because the big difference between processor speed and main memory access speed. Thus, a cache miss is still relatively expensive in main memory databases.
Accordingly, we will see that many of the index data structures proposed for main memory databases focused on being cache conscious or cache sensitive. In
summary, a main memory index should be able to reduce overall computation time while using as little memory as possible~\cite{lehman1986study}. It should also
be noted that unlike disk based systems where it is advantageous to store actual attribute values in the index, main memory systems do not require that.
Instead, pointers to the actual attribute values can be used. This, in turn, will reduce the size of the index which is preferable in this case.

In this section, we talk about three families of index structures used in MMDB: B Trees, T Trees, and Binary Search Trees. There are more recent index
structures developed for MMDB such as J+ Trees~\cite{luan2009prefetching}, but which have not gained much popularity, and so we do not discuss them in our
survey. %Table~\ref{tab:mmdbindexsumm} summarizes the features of the different index structures discussed in this section with respect to MMDB. The T Tree is
the most used index for MMDB (E.g. MySQL CLuster). However, there are other indexes used as well. For example, IBM's solidDB uses Trie (or prefix tree) for its
indexing. For their purposes, it served as a good index for main memory since it eliminates the need for many key comparisons~\cite{ibmsoliddb}.

 \begin{table}[t!]
\centering
\scalebox{0.6}{
\begin{tabular}{p{2cm}|p{6cm}|p{6cm}}
\hline
 \textbf{Index Structure} & \textbf{Features} & \textbf{Cache Consciousness}\\
\hline\hline
B Trees &
\begin{itemize}\item Good storage utilization
\item Reasonably quick searching
\item Fast updating
\end{itemize} 
& \begin{itemize}\item Reasonable cache behavior if node fits in cache line\end{itemize}\\
\hline
B+ Trees & 
\begin{itemize}
\item Reasonably quick searching
\item Fast updating
\end{itemize} &
\begin{itemize}
\item Reasonable cache behavior if node fits in cache line\end{itemize}\\
\hline
CSB+ Trees & \begin{itemize}
              \item Same features of B+ Trees
\item Stores child nodes in an array. Stores pointer to first child only and uses index calculation to get the rest
             \end{itemize} &
\begin{itemize}
 \item Good cache performance due to improved locality \& lower tree height (because of larger node size)
\end{itemize}\\
\hline

pB+ Trees & \begin{itemize}
              \item Same features of B+ Trees
             \end{itemize} &
\begin{itemize}
 \item Improves cache miss performance by prefetching more data in each cache miss
\item Node size is also increased leading to the same benefits of CSB+ Trees
\end{itemize}\\
\hline
T Tree &  \begin{itemize}
              \item Contain pointers to data values instead of the values themselves leading to better space utilization
\item Many sorted keys per node
\item Only minimum and maximum keys used for comparisons
\item Contains record pointers in every key
\item Good update \& storage characteristics if Btrees.

             \end{itemize} &
\begin{itemize}
 \item Cache behavior was not considered at time of design
\item Record pointers waste space in cache
\end{itemize}\\
\hline
CST & \begin{itemize}
       \item Binary search tree for maximum value of each node leads to faster search
\item More space efficient by removing pointers \& using array storage. Index calculation used to allocate values
      \end{itemize} &
\begin{itemize}
 \item Node size is aligned with cache line size to avoid misses
\item Higher usage of cached data in binary search tree
\end{itemize}\\
\hline
CSS Trees &  \begin{itemize}
       \item Fast traversal in $log_{m+1}n$ ($m$ is the number of keys per node)
\item Mainly suitable for read environments
      \end{itemize} &
\begin{itemize}
 \item Good cache behavior since $m$ is chosen to fit in the cache line size
\end{itemize}\\

@@

Although T Trees have been specifically designed for MMDB, they did not consider the cache behavior at that time since the gap between processor speed and main
memory speed was not big then~\cite{rao1999cache}. Although many keys are
fit into one node, only the two end keys are used for comparison leading to low node utilization. Similarly, array binary search trees have the same problem. 
It was discovered that B+-trees outperform
T-Trees on modern processors because of the growth of cache miss latency relative to processor speed~\cite{rao1999cache,lee2007cst}. This is because the height
of the tree is high which makes the total number of memory accesses from the root to the leaf node higher. The other reason is that the node size is not aligned
with the cache line size. Another problem with T-Trees is that there is a big waste of space in the cache because unnecessary data which is not used (the
record pointers) brought into the cache.

To resolve these issues, Lee et al.~\cite{lee2007cst} propose the Cache Sensitive T-Trees (CST-Trees). First, a binary
search tree whose node values are the maximum key of each node in the T Tree is constructed. That way, for a given value, the node containing this value could
be quickly located through searching the binary tree. Second, the need to store pointers is eliminated by storing the tree in an array and locating the
necessary nodes through index calculation. This produces a better utilization of the cache by removing the unused data. Finally, node sizes are aligned with
cache line size such that there are no cache misses when accessing data in each
binary search tree in the array.

@@

Binary Search Trees have poor cache behavior when the array is much bigger than the cache. Rao and Ross~\cite{rao1999cache} propose the Cache Sensitive Search
Tree (CSS-Tree) to improve this cache behavior. To do that, the search tree is stored as a directory in an array. The number of keys per node is chosen such
that the whole node fits in a cache line. This improves local searching within a node to just one cache miss. They also hard code the traversal within a node
such that finding the next node happens in $log_{m+1}n$ times rather than $log_{2}n$ times in a binary search tree where $m$ is the number of keys per node.
However, CSS Trees are only suitable for read environments since they eliminate nearly all child pointers, thus removing the support for incremental update.

\subsubsection{Materialized Views}
From our literature survey, it seems that materialized views are not used in MMDB. This makes sense since the problems materialized views try to address are
not present in MMDB. That is, the cost of computing complicated join queries is much less in MMDB due to the low cost of accessing the data in memory.
Accordingly, the maintenance cost of materialized views in MMDB will outweigh their minimal benefit. However, this does not mean that they cannot be used at
all in MMDBs. As with all other physical design decisions, it rather depend on the benefits of implementing the technique. Accordingly, if the data is very
large and may benefit from using materialized views, there is no reason not to use them.

@@

Partitioning is used in disk based systems to divide the data into the physical pages in a way that will yield the best performance. This could be through
vertical partitioning or horizontal partitioning, as previously explained. In main memory databases, we do not have the issue of expensive disk access which
makes partitioning in memory unnecessary as it will not save any costs. However, partitioning is needed in the secondary storage of the data on
disk~\cite{gruenwald1990database}. Disk storage is still used as a backup of main memory databases. When a crash occurs, a reload of the database from disk or
archive memory takes place. In order to avoid page faults caused by referencing data that is still being reloaded, the reloading process must be designed to be
efficient in loading the important data such that the database users are not affected. An efficient MMDB reload takes into consideration the number of I/Os for
reload and the number of main memory references during transaction processing. Gruenwald and Eich~\cite{gruenwald1990database,gruenwald1990choosing} show that
horizontal partitioning is the most suitable when the database performs more modifications than tuple deletions and more selections that projections and joins.
Vertical partitioning is chosen in the other cases. They also show that if we just concentrate on the reload performance (i.e the number of I/Os for
reload), then vertical partitioning is always the best choice.

@@

XML (eXtensible Markup Language) is increasingly being used as a data exchange format. Accordingly, the ability to store and manage XML documents is needed.
Salminen and Tompa~\cite{salminen2001requirements} define an XML database as ``a collection of XML documents and their parts, maintained by a system
having capabilities to manage and control the collection itself and the information represented by that collection.'' Salminena and Tompa also highlight some of
the requirements of an XML database system. These requirements include querying parts of an XML document, and
transformations of XML documents. The main challenge in XML databases is path navigation since it is a very expensive operation.

There are two ways in which an XML database system can be designed. The first is to have the database internally translate the XML document into relational
tables (XML-enabled databases). The second is to have data structures that can persistently store XML documents (native XML databases). XML-enabled databases
are suited for data-centric documents which have a regular structure without any mixed content~\cite{bourret2003xml}. On the other hand, for document-centric
documents which have an irregular structure with lots of mixed content, having a native XML database is essential. Examples of XML-enabled databases include SQL
Server 2005 and Oracle, and examples of native XML databases include Tamino~\cite{taminodb},
eXist~\cite{meier2009exist}, and TIMBER~\cite{jagadish2002timber}.

In this section, we will discuss how the database's physical design needs to be adapted in both cases. Other challenges such as expressing and executing queries
and updates are faced in XML database design, but these are beyond the scope of this survey.

@@

Despite the special nature of XML documents, the same index structures used in relational databases can still be utilized in XML databases. However, some
adjustments may need to be made to fit the nature of the XML documents. For XML-enabled databases, XML documents are usually represented as relational tables,
and then indexed similar to other tables. Pal et al.~\cite{pal2004indexing} describe how this is done in Microsoft SQL Server 2005. To start with, a new data
type called `XML' was introduced. This data type could contain values of complete XML documents, or just fragments of XML data. First, the different nodes in
the XML data are labeled with the ORDPATH~\cite{o2004ordpaths} mechanism. The XML data is then shredded into a relational table with five main columns: ORDPATH,
TAG, NODE\_TYPE, VALUE, and PATH\_ID. This information is then stored in a $B^{+}-tree$ as the primary index of the XML type. Additionally, a secondary index
can be created on any of the columns of the primary index to speed things up.

For native XML databases such as eXist~\cite{meier2009exist} or TIMBER~\cite{jagadish2002timber}, B+trees are still used to index the XML documents.
However, in contrast to XML-enabled databases, the XML data is not first stripped into a relational table. eXist uses a number schema to assign unique
identifiers to each node in the XML document. These unique number identifiers allow the determination of any node's parent, sibling or possible child nodes.
Four index files are created for XML data where all the indexes are based on $B^{+}-trees$. One index manages the collection hierarchy, one index collects nodes
in a paged file and associates unique node identifiers to the actual nodes, one index indexes elements and attributes, and the last index keeps track of word
occurrences and is used by the full text search extensions. Oracle uses a similar technique for indexing its XML content where an XMLIndex table is created for
XML entries~\cite{oracleXML}. This table contains a unique identifier for the XPath leading to the node, the row id of the table used to store the XML data,an
order key that identifies the hierarchal positioning of the node, a locater key used for fragment extraction, and the text value of the node.

TIMBER also uses a similar numbering schema. There are minor differences between the two numbering techniques, but the main idea is the same. The numbering is
based on three values of a node: its start label, its end label, and its level (i.e. nested depth). On the other hand, Sedna~\cite{taranov2010sedna} uses a
different reasoning to produce its numbering schema. Sedna assigns string identifiers to nodes such that they are lexically ordered according to the position
of the node. Despite the differences in the ways the numbering is constructed, it seems that existing indexing techniques can be adapted for XML databases. The
challenge is how to assign the nodes unique identifiers (the numbers or labels in the numbering schema) that can be used as identifiers in the index to
reflect the XML document structure. Additionally, in native XML databases, documents are usually grouped into collections, and an index may be used to search
within these collections. This is different from the node index which searches within a document.
